Allee Effect and Cultural Evolution

### 2015-03-25 manuscript/draft.tex
majority of equations defined

\usepackage[utf8]{inputenc}
\usepackage{xcolor}

@@

\section{The Model}
We consider the evolutionary dynamics of two mutually exclusive subsistence strategies which fitness have both positive and negative frequency dependence. In particular we examine how two population of individuals, m and n, change over time if we allow for a social learning strategy guided by the difference in the fitness of the individuals. The model can theoretically be extended to more than two strategies, but here we decide to explore a two trait version for simplicity. The core model can be depicted with the following coupled difference equation:

\xavi{equation1}

where the population of each trait at time t+1 is the given by the number of individuals during the previous time-step (mt and nt), the number of offspring generated by each population (R(m,t) and R(n,t)) and the number of individuals shifting from one strategy to the other (C(m,n,t)).

The intrinsic growth rate of each population can be obtained from the following paired equation:

\xavi{equation2}

where r is the basic growth rate, A is the allee threshold, and K is the carrying capacity, with A<K.  The intrinsic growth rate will thus become negative when the population size is below A or above K, and will reach its maximum value Rmax at (K-A)/2, with Rmax equivalent to r(A-K)2/4AK.

Cultural transmission is portrayed here as a migration from the population with lower payoff to the population with higher payoff:

\xavi{equation3}

where delta defines the proportion of individuals shifting strategy. This is given by the following equation:

\xavi{equation4}

where z is the maximum transmission rate (i.e. the highest proportion of individuals shifting strategy), DELTA XX is the maximum between Rmax(m,t) and Rmax(n,t) , and b is a calibration parameter that measure the perception of the difference in the payoffs (i.e. small values of b determines a higher rate by which the theoretical maximum transmission rate is reached).

@@

We investigate different initial conditions of m and n for 6 different values of z (0,0.05,0.1,0.2,0.3, and 0.5). For each parameter combination we apply equations 1~4 for 50,000 iterations and record the final values of m and n, identifying the following equilibria

@@

\xavi{equation5}

@@

The first scenario assumes two traits that are mathematically equivalent, so that  Am = An
, Km=Kn and rm=rn. Figure 1 shows the basin of attractions of each of the four equilibria for different values of z, setting r=0.005, .A=0.2 and K=1. When the z=0, i.e. when cultural transmission is disabled we have straight boundaries between the basins of E1, E2m, E2n, and E3 defined by Allee threshold. In other words when both one or both the population are below A we have some extinction events (E3 when mt=0<Am and nt=0<An, E2m when mt=0>Am and nt=0<An and E3n when mt=0<Am and nt=0>An). When z>0, the boundaries of these basins changes and with high z we see the emergence of unstable equilibria (U).

@@

The model described in equations [1]~[4] assumes that the resource pool of the two strategy are completely independent to each other. In reality many strategies have some degree of competition, whereby the frequency dependence is not exclusively internal. In other words, the frequency of one  trait might increase or decrease the fitness of the other. The following equation can substitute equation [2]:

>>>>

\usepackage[utf8]{inputenc}
\usepackage{xcolor}
\usepackage{amsmath}

@@

\section{The Model}
We consider the evolutionary dynamics of two mutually exclusive subsistence strategies which fitness have both positive and negative frequency dependence. In particular we examine how two population of individuals, m and n, change over time if we allow for a social learning strategy guided by the difference in the fitness of the individuals. The model can theoretically be extended to more than two strategies, but here we decide to explore a two trait version for simplicity. The core model can be depicted with the coupled difference equation \eqref{eq1}:

\begin{align}
\label{eq1}
\begin{cases}
m_{t+1}& = m_t + m_t R_{m,t} + C_{m,n,t} \\
n_{t+1}& = n_t + n_t R_{n,t} + C_{m,n,t}
\end{cases}
\end{align}

where $m_{t+1}$ and $n_{t+1}$ is the population of each trait at time $t+1$ is the given by the number of individuals during the previous time-step ($m_t$ and $n_t$), $R_{m,t}$ and $R_{n,t}$ is the number of offspring generated by each population and $C_{m,n,t}$ is the number of individuals shifting from one strategy to the other.

The intrinsic growth rate of each population can be obtained from the paired equation \eqref{eq2}:

\begin{align}
\label{eq2}
\begin{cases}
R_{m,t}& = r_m (\frac{m_t}{A_m}-1)(1-\frac{m_t}{K_m})\\
R_{n,t}& = r_n (\frac{n_t}{A_n}-1)(1-\frac{n_t}{K_n})
\end{cases}
\end{align}

where $r$ is the basic growth rate, $A$ is the allee threshold, and $K$ is the carrying capacity, with $A < K$.  The intrinsic growth rate will thus become negative when the population size is below $A$ or above $K$, and will reach its maximum value $R_{max} = \frac{(K-A)}{2}$, with $R_{max}$ equivalent to $\frac{r(A-K)2}{4AK}$. 
\xavi{I think this last sentence is confusing}

Cultural transmission is portrayed here as a migration from the population with lower payoff to the population with higher payoff, as seen in \eqref{eq3}:

\begin{align}
\label{eq3}
C_{m,n,t} = 
\begin{cases}
0& \text{if } R_{m,t} = R_{n,t}\\
\zeta(m_t+m_tR_{m,t})& \text{if } R_{m,t} > R_{n,t}\\
-\zeta(n_t+n_tR_{n,t})& \text{if } R_{m,t} < R_{n,t}
\end{cases}
\end{align}

where $\zeta$ defines the proportion of individuals shifting strategy, and given by \eqref{eq4}:

\begin{align}
\label{eq4}
\zeta = 
\begin{cases}
z& \text{if }|R_{m,t}-R_{n,t}| > b\Delta\\
z\frac{|R_{m,t}-R{n,t}|}{b\Delta}& \text{if }|R_{m,t}-R_{n,t}| \leq b\Delta\\
\end{cases}
\end{align}

where $\zeta$ is the maximum transmission rate (i.e. the highest proportion of individuals shifting strategy), $\Delta$ is the maximum between $R_{max(m,t)}$ and $R_{max(n,t)}$, and $b$ is a calibration parameter that measure the perception of the difference in the payoffs (i.e. small values of b determines a higher rate by which the theoretical maximum transmission rate is reached).

@@

We investigate different initial conditions of $m$ and $n$ for 6 different values of z (0,0.05,0.1,0.2,0.3, and 0.5). For each parameter combination we apply equations (\ref{eq1},~\ref{eq4}) for 50,000 iterations and record the final values of $m$ and $n$, identifying the following equilibria.

@@

\xavi{equation5, lyapunov, where this does go?}

@@

The first scenario assumes two traits that are mathematically equivalent, so that $A_m = A_m, K_m=K_n and r_m=r_n$. Figure 1 shows the basin of attractions of each of the four equilibria for different values of z, setting r=0.005, .A=0.2 and K=1. When the z=0, i.e. when cultural transmission is disabled we have straight boundaries between the basins of E1, E2m, E2n, and E3 defined by Allee threshold. In other words when both one or both the population are below A we have some extinction events (E3 when mt=0<Am and nt=0<An, E2m when mt=0>Am and nt=0<An and E3n when mt=0<Am and nt=0>An). When z>0, the boundaries of these basins changes and with high z we see the emergence of unstable equilibria (U).

@@

The model described in equations  (\ref{eq1},~\ref{eq4}) assumes that the resource pool of the two strategy are completely independent to each other. In reality many strategies have some degree of competition, whereby the frequency dependence is not exclusively internal. In other words, the frequency of one  trait might increase or decrease the fitness of the other. The following equation can substitute equation [2]:
